---
title: "Introduction to data.table"
author: "Kim Eng Ky <br/> @kykimeng <br/> kim_ky[at]uhc[dot]com <br/> kykimeng.com/slides/rnorth_data_table.html"
date: "August 16, 2019"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "#000099",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "500", "500i"),
  code_font_google   = google_font("Droid Mono"),
  text_font_size = '24px',
  code_highlight_color = "#FFFFFF"
)
```

# What is `data.table`?

"`data.table` is an R package that provides an enhanced version of `data.frames`." 

- faster
- memory efficient 
- concise syntax
- an excellent addition to your R venacular, particularly if you work with massive datasets.

---

# Install and load

```{r, eval=F}
# install 
install.packages("data.table")
# load 
library(data.table)
```


---


# General form

```{r,echo=F, fig.width=10, fig.height=5}
library(data.table)
library(ggplot2)
ggplot() + 
  geom_text(aes(x = 0, y = 0, label = "DT[i, j, by]"), size = 16) +
  theme_classic() +
  theme(axis.line = element_blank(), axis.title = element_blank(), 
        axis.text = element_blank(), axis.ticks = element_blank()) +
  geom_line(aes(x = c(-2, -.3), y = c(-2, -0.25)), arrow = arrow(), size = 1) +
  geom_line(aes(x = c(0.15, 0.15), y = c(-2, -0.25)), arrow = arrow(), size = 1) +
  geom_line(aes(x = c(.9, 2), y = c(-.25, -2)), arrow = arrow(ends = "first"), size = 1) +
  geom_label(aes(x = c(-2.5, 0.15, 3), y = -2, label = c("WHERE\n(on which rows)", "SELECT\n(do what)", "GROUP BY\n(group by what)")), size = 10) +
  scale_x_continuous(limits = c(-4, 5)) +
  scale_y_continuous(limits = c(-2.2, .5))
```

---

A lot of `data.table` commands are similar to the usual R `data.frame`.

```{r, eval=F}
# create a data.table
DT <- data.table(x = c(1, 2, 3), 
                 y = c('a', 'b', 'c'))

# turn airquality data.frame to 
# a data.table object
data("airquality")
airquality_dt <- as.data.table(airquality)

# turn iris into a data.table 
# instead of making a copy
setDT(iris)
```

```{r, include=F}
data("airquality")
airquality_dt <- as.data.table(airquality)
```

---

When there are 100+ rows, `data.table` only prints the first five and the last five rows.

```{r}
airquality_dt
```


---

# Subset rows

```{r, eval=F}
# data.frame
airquality[1:3,]
airquality[airquality$Month == 5,]
airquality[airquality$Month == 5 & 
             airquality$Ozone > 100,]

# data.table
airquality_dt[1:3]
airquality_dt[Month == 5]
airquality_dt[Month == 5 & Ozone > 100]
```

---

# Subset columns

```{r, eval=F}
# data.frame
airquality[, c('Month', 'Day', 'Ozone')]

# data.table
airquality_dt[, .(Month, Day, Ozone)]
```


---

# Manipulation on columns

Calculate mean Ozone value:

```{r}
# data.frame
mean(airquality$Ozone, na.rm = T)

# data.table
airquality_dt[, mean(Ozone, na.rm = T)]
```

---

Calculate mean Ozone value by Month

```{r, eval=F}
# data.frame 
aggregate(Ozone ~ Month, airquality, 
          mean, na.rm = T)
```

```{r}
# data.table
airquality_dt[, mean(Ozone, na.rm = T), 
              by = Month]
```

---

Calculate mean Ozone value when Temp is above 70 by Month

```{r, eval=F}
# data.frame 
aggregate(Ozone ~ Month, 
          airquality[airquality$Temp > 70,], 
          mean, na.rm = T)
```

```{r}
# data.table
airquality_dt[Temp > 70, 
              mean(Ozone, na.rm = T), 
              by = Month]
```

---

With `data.table`, you can also use expressions within `by` statement.

```{r}
airquality_dt[, mean(Ozone, na.rm = T),
              by = .(Temp > 70)]
```

Names output columns to be more meaningful:

```{r}
airquality_dt[, .(
  mean_ozone = mean(Ozone, na.rm = T)
), by = .(temp_70_plus = Temp > 70)]
```

---

In `data.table`, there are a few very useful shortcut commands. `.N` is a counting tool that holds the number of observations in the current group.

```{r}
# total number of rows
airquality_dt[, .N]
```

Or we can calculate number of rows by month:

```{r}
airquality_dt[, .N, Month]
```

---

`uniqueN()` calculates number of unique values. 

```{r}
# number of different months in data
airquality_dt[, uniqueN(Month)]
```

Or we can calculate number of days in each month.

```{r}
airquality_dt[, uniqueN(Day), Month]
```



---

`.GRP` command allows you to add unique id to groups as identified by `by` statement.

```{r}
# just one group
airquality_dt[, .GRP]

# group id by month
airquality_dt[, .GRP, Month]
```

---

Or a bit more complicated

```{r}
airquality_dt[, .GRP, .(
  Month, 
  temp_70_plus = Temp > 70)]
```



--- 
---

And, `.I` holds the row numbers.

```{r}
airquality_dt[, .I]
```


---


# Keys

`data.table`s are ordered by its keys in increaing order. Keys used to be required for "keyed" joins. Now we can do "ad-hoc" joins using `on=` without having to do `setkey()` prior to joining.

---

# Left/Right Joins

```{r, eval=F}
sunny_days <- data.table(
  Month = c(5, 6, 7, 8, 9),
  sunny_days = c(20, 25, 28, 23, 15), 
  key = 'Month'
)
# data.frame
merge(airquality, sunny_days, 
      by = 'Month', all.x = F, all.y = T)
# keyed join
setkey(airquality_dt, Month)
airquality_dt[sunny_days]
# ad-hoc join
airquality_dt[sunny_days, on = .(Month)]
```



---

We can also filter using joins. 

```{r}
(filter_vals <- data.table(
  Month = rep(c(5, 6), each  = 3),
  Day = rep(c(1, 2, 3), 2)
))
```



---



Subset `airquality` based on values in `filter_vals`. 

```{r, eval=F}
# only keep rows that match 
# Month/Day pair in filter_vals
airquality_dt[filter_vals, 
              on = .(Month, Day)]
```

```{r}
# or 
setkey(airquality_dt, Month, Day)
setkey(filter_vals, Month, Day)
airquality_dt[filter_vals]
```

---

What if we want to exclude rows that match the Month/Day pair in `filter_vals`?

```{r, eval=F}
airquality_dt[!filter_vals, 
              on = .(Month, Day)]
```

```{r}
# or after setting keys:
airquality_dt[!filter_vals]
```


---

# Inner Joins

```{r, eval=F}
# data.frame 
merge(airquality, sunny_days, 
      by = 'Month')

# data.table
airquality_dt[sunny_days, 
              on = .(Month),
              nomatch = 0]
```

---

# Full Joins

```{r, eval=F}
merge(airquality_dt, sunny_days, 
      by = 'Month', all = T)
```

---

# Defining Function `:=`

You can define new variables in `data.table` with the `:=`.

```{r}
# create a new variable called high_temp
airquality_dt[, high_temp := Temp > 70]
head(airquality_dt, 3)
```

---

Or create a new variable to hold mean values of wind by month. 

```{r, include=F}
airquality_dt[, high_temp := NULL]
```

```{r}
airquality_dt[, mean_wind := mean(Wind),
              by = Month]
head(airquality_dt, 3)
```

---

We can also create multiple variables in one go.

```{r, eval=F}
airquality_dt[, `:=` (
  high_temp = Temp > 70, 
  row_id = .I
)]
# or equivalently
airquality_dt[
  , c('high_temp', 'row_id') 
      := list(Temp > 70, .I)
]
```

---

One great benefit of `data.table` is the ability to sub-assign by reference.

```{r, eval=F}
airquality_dt[high_temp == TRUE, 
              new_wind := Wind]
```

---

We can also use `:=` to remove columns.

```{r, eval=F}
airquality_dt[, high_temp := NULL]
# or multiple ones at a time
airquality_dt[
  , c('high_temp', 'mean_wind') := NULL
]
# or 
cols <- c('high_temp', 'mean_wind')
airquality_dt[, (cols) := NULL]
```

---

# `.SD` and `.SDcols`

You can access columns of a `data.table` using `.SD` syntax.

Select all columns in `airquality_dt`:

```{r, include=F}
airquality_dt[, c('mean_wind', 'row_id', 'high_temp') := NULL]
```

```{r}
head(airquality_dt[, .SD], 3)
```

---

Select subset of columns by specifying columns in `.SDcols`:

```{r}
head(airquality_dt[
  , .SD, .SDcols = c('Month', 'Day')
], 3)
# Or, use column indices.
head(airquality_dt[
  , .SD, .SDcols = c(5, 6)
], 3)
```

---

Round up `Ozone`, `Solar.R` and `Wind` variables. 

```{r, eval=F}
airquality_dt[, lapply(.SD, round), 
              .SDcols = c('Ozone', 'Solar.R', 'Wind')]
```

We can overwrite them into the original variables.

```{r, eval=F}
cols <- c('Ozone', 'Solar.R', 'Wind')
airquality_dt[, (cols) := lapply(.SD, round), 
              .SDcols = cols]
```

--- 
---

Calculate mean by `Month`:

```{r}
airquality_dt[, lapply(.SD, mean, na.rm = T), 
              .SDcols = 1:4, 
              by = Month]
```

---

Select first and last rows:

```{r}
airquality_dt[, .SD[c(1, .N)]]
```

---

Select first and last rows for each month:

```{r}
airquality_dt[, .SD[c(1, .N)], 
              by = Month]
```


---

# Chaining

Chaining in `data.table` refers to using multiple `[]` in one command.

Calculate mean `Temp` by `Month`, order by the calculated mean, and only keep rows with mean less than 70:

```{r}
airquality_dt[, .(
  mean_temp = mean(Temp, na.rm = T)
), by = Month][
  order(mean_temp)
][
  mean_temp < 70
]
```


---

# Rolling Joins

Let’s **roll forward** (last observation carried forward):

```{r}
dt1 <- data.table(
  x = c("a", "b", "c", "d"), 
  dt1_y = c(11.9, 21.4, 5.7, 18)
)
dt2 <- data.table(
  dt2_y = c(10, 15, 20), 
  z = c("one", "two", "three")
)
# duplicate y  
dt1[, `:=` (joint_y = dt1_y)]
dt2[, `:=` (joint_y = dt2_y)]
```

---

```{r}
dt1
dt2
```

---

```{r}
dt2[dt1, on = .(joint_y), roll = T] 
# or: dt2[dt1, on = .(joint_y), roll = Inf]
```

This means that for each `dt1_y`, find `dt2_y` that is closest to `dt1_y` with the condition `dt1_y >= dt2_y`. This condition is the forward part of the rolling join.

---

Now, let’s **roll backward** (next observation carried backward):

```{r}
dt2[dt1, on = .(joint_y), roll = -Inf]
```

Now the condition is reverse, i.e. `dt1_y <= dt2_y`. Pay attention to the minus sign in the `-Inf`; it is how you tell data.table to roll backward instead of forward.

---

Next, let’s give them a window to roll on:

```{r}
dt2[dt1, on = .(joint_y), roll = -2]
```

Now the condition is restricted to only match if `dt1_y <= dt2_y - 2`.

---

Or, we can roll with window in the other direction.

```{r}
dt2[dt1, on = .(joint_y), roll = 2]
```

The condition here is `dt1_y <= dt2_y + 2`.

---

Or, we can roll both ways to the nearest value.

```{r}
dt2[dt1, on = .(joint_y), roll = "nearest"]
```

---

Another cool feature in `data.table` is the `foverlaps()` function which allows you to join tables based on range of values. 

Each window does not have to be equal.

```{r}
(dt3 <- data.table(
  min_y = c(0, 10, 15, 20), 
  max_y = c(10, 15, 20, 30),
  key = c('min_y', 'max_y')
))
```

---

```{r}
(dt4 <- data.table(
  x = c('a', 'b', 'c', 'd'),
  dt4_y = c(11.9, 21.4, 5.7, 18),
  dt4_y_end = c(13, 25, 10, 22), 
  key = c('dt4_y', 'dt4_y_end')
))
```


---

```{r}
foverlaps(dt4, dt3, type = "any")
```

When specifying `type = "any"`, as long as `[dt4_y, dt4_y_end]` and `[min_y, max_y]` ranges overlap it will give a match.

---

```{r}
foverlaps(dt4, dt3, type = "within")
```

When specifying `type = "within"`, the only matches returned are those where `[dt1_y, dt1_y_end]` range is within `[min_y, max_y]`.

---

# References

1. Matt Dowle and Arun Srinivasan (2018). `data.table`: Extension of `data.frame`. R
  package version 1.11.4. https://CRAN.R-project.org/package=data.table
2. https://github.com/Rdatatable/data.table/wiki
3. `data.table` vs. `dplyr`. https://stackoverflow.com/questions/21435339/data-table-vs-dplyr-can-one-do-something-well-the-other-cant-or-does-poorly
4. Advaned tips and tricks with `data.table`. http://brooksandrew.github.io/simpleblog/articles/advanced-data-table/

---

# More references

5. `data.table` cheat sheet. https://s3.amazonaws.com/assets.datacamp.com/img/blog/data+table+cheat+sheet.pdf
6. Purpose of setting a key in `data.table`. https://stackoverflow.com/questions/20039335/what-is-the-purpose-of-setting-a-key-in-data-table
7. `.EACHI` in `data.table`. https://stackoverflow.com/questions/27004002/eachi-in-data-table/27004566#27004566
8. Understand `data.table` rolling joins. https://www.r-bloggers.com/understanding-data-table-rolling-joins/