---
title: "Brilliant Insights"
author: "Matt Pettis"
date: "`r Sys.time()`"
output:
    html_document:
        theme: journal
        highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# My Thesis

I can tell the difference between any of the different iris species!



# My Analysis

## What does my dataset look like?

```{r}
print(here::here())
print(iris)
```

Oooh too much data to understand all at once.  Let's take the means of all of
the dimensions for each type of iris:

```{r}
df_iris <- dplyr::group_by(iris, Species)
df_iris_summary <- dplyr::summarise_all(df_iris, mean)

# Display table nicely
knitr::kable(df_iris_summary)
```

> Wow!  It looks like setosa has different Petal dimensions!
> Well, at least according to the mean values.  Wouldn't it be nice if we could
just see the points and see if anything pops?





# We can see the points!  Here we plot it!

```{r}
ggplot2::ggplot(iris) +
    ggplot2::geom_point(ggplot2::aes(Petal.Width, Petal.Length, color=Species)) +
    ggplot2::ggtitle("Holy Smokes Would You Look at This!")
```

Good grief!  Not only did the setosas look really different, but it looks like:

- you can distinguish between the versicolors and virginicas too!
- We should look at some pictures and see if this makes sense:



# Pictures of these irises

Setosa:

![setosa](`r file.path(here::here(), "img", "setosa.jpg")`)

Versicolor:

![versicolor](`r file.path(here::here(), "img", "versicolor.jpg")`)

Virginica:

![virginica](`r file.path(here::here(), "img", "virginica.jpg")`)

> It's... obvious?
